---
title: "Length to Lipids conversion.V2"
author: "P. Santiago Domínguez-Sánchez"
date: "2025"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(readr)
library(kableExtra)
library(zoo)
library(vctrs)
library(here)
```

### Calculating lengths at 'RMIS region - Brood year - Age' level

```{r data sets}
length_ref_val <- read_xlsx('/Users/domingpa/Documents/Github Repositories/Chinook_energetics/R/output/GAM_reference_values.xlsx')
head(length_ref_val)

gam_predictions  <- read_xlsx('/Users/domingpa/Documents/Github Repositories/Chinook_energetics/R/output/GAM_predictions.xlsx')
head(gam_predictions)

colnames(gam_predictions)[2:3] <- c("ocean_age","brood_year")

```


### Size at age and region

```{r wider table, echo = FALSE}
w_size_at_age <- gam_predictions%>%select(1:4)%>%pivot_wider(names_from = ocean_age, values_from = fit)

w_size_at_age[29:39,]%>%
  mutate_all(~ ifelse(is.na(.), cell_spec("NA", "html", background = "red"), .)) %>%
  kable(escape = FALSE, format = "html") %>%
  kable_styling(full_width = F)

```

In this case, the NAs belonging to the CECR 2008-2010 and FRTH 1980 - 1987 at age 5 are due to the 'rel_pred' and 'reference size' value being missing at this age.

Let's check where the NAs are

```{r check NAs, echo = FALSE, warning=FALSE, message=FALSE}
check.grid <- expand.grid(region = unique(gam_predictions$region), brood_year = unique(gam_predictions$brood_year), ocean_age = unique(gam_predictions$ocean_age))

NA_size_at_age <- gam_predictions%>%
  full_join(check.grid,gam_predictions, by = c("region","brood_year","ocean_age"))%>%
  filter(is.na(fit))%>%
  group_by(region,ocean_age)%>%
  summarise()

```

```{r NAs size at age}

print(NA_size_at_age)
```

All regions have at least one missing value at 'age-5', and only the 'GRAY' region has missing values at ages 1 and 5. Since the FRAM abundance database does not contain information for age 1, the missing values at 'age-1' will not be relevant in the next calculations.

But before filling the NAs, let's see how the data looks so far.

```{r prod size pred, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=3}
avg_size_at_year <- gam_predictions%>%
  group_by(ocean_age,brood_year)%>%
  summarise(avg.size = mean(fit, na.rm = TRUE))%>%
  ungroup()

ggplot(gam_predictions)+
  geom_line(mapping = aes(x = brood_year, y = fit, color = region))+
  geom_line(mapping = aes(x = brood_year, y = avg.size, group = ocean_age), data = avg_size_at_year, color = "black", size = 1)+
  facet_wrap(~ocean_age,nrow = 1,ncol = 5, scales = "free")+
  labs(x = "Brood year", y = "Length (mm)")+
  theme_light()

```
To fill the NAs, **I tackled this issue based on two assumptions:** 

**1)** If there was data available for age 5, I kept the 'fitted lengths', 'se fitted lengths', 'reference means' and "rel_pred" constant for the previous years relative to the last year with information.

**2)** For the years for age 5 without the 'fitted lengths', 'se fitted lengths', 'reference means' and "rel_pred", the values were taken relative to age 4.

Additionally, JUAN-age 5 is removed since the size changes were estimated from few samples, resulting in a dramatic size decrease. The age-5 sizes were calculated based nder assumption # 2. 
```{r fill NAs, warning=FALSE, message=FALSE, fig.width=12, fig.height=3}

gam_predictions <- gam_predictions%>%
  filter(!(region == 'JUAN' & ocean_age == 5))

full_gam_predictions <- check.grid%>%
  left_join(gam_predictions, by = c("region","brood_year","ocean_age"))%>%
  group_by(region, ocean_age)%>%
  arrange(region,ocean_age)%>%
  fill(fit, se.fit, ref_mean, rel_pred, .direction = 'down')%>% #Fill NAs under assumption 1
  ungroup()%>%
  group_by(region, brood_year)%>%
  arrange(region,brood_year)%>%
  fill(fit, se.fit, ref_mean, rel_pred, .direction = 'down')%>% #Fill NAs under assumption 2
  ungroup()

avg_size_at_year <- full_gam_predictions%>%
  group_by(ocean_age,brood_year)%>%
  summarise(avg.size = mean(fit, na.rm = TRUE))%>%
  ungroup()
```

*Chinook salmon size over brood year  - Each line is a region and color indicates ocean age*
```{r fill NAs plot 1, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4, fig.height=6}
ggplot(full_gam_predictions, aes(x = brood_year, y = fit, color = ocean_age, group = interaction(region, ocean_age)))+
  geom_line(alpha = 0.5)+
  scale_color_viridis_c(option = 'plasma', direction = -1)+
  labs(x = 'Brood year', y = 'Length (cm)')+
  theme_light()
```

*Chinook salmon size changes divided by ocean age - Color indicates region and black line is the mean size trend for each age*
```{r fill NAs plot 2, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=3}
ggplot(full_gam_predictions)+
  geom_line(mapping = aes(x = brood_year, y = fit, color = region))+
  geom_line(mapping = aes(x = brood_year, y = avg.size, group = ocean_age), data = avg_size_at_year, color = "black", size = 1)+
  facet_wrap(~ocean_age,nrow = 1,ncol = 5)+#, scales = "free")+
  labs(x = "Brood year", y = "Length (mm)")+
  theme_light()
```

*Size changes over brood year for each region - Color indicates ocean age*
```{r fill NAs plot 3, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=6}
fit_by_regions <-ggplot(full_gam_predictions)+
  geom_line(mapping = aes(x = brood_year, y = fit, color = as.factor(ocean_age)))+
  facet_wrap(~region,nrow = 3,ncol = 7)+#, scales = "free")+
  labs(x = "Brood year", y = "Length (mm)")+
  theme_minimal()

fit_by_regions
```

### Add Lipid tier to FRAM abundance database

I manually organized this database in Excel. [O'neill et al. 2014](https://www.int-res.com/articles/esr_oa/n025p265.pdf), calculated the lipid tier for each stock at the FRAM stock level, and [Ohlberger et al. 2018](https://onlinelibrary.wiley.com/doi/full/10.1111/faf.12272) estimated the salmon stock lengths at the RMIS level. Hence, I matched FRAM stocks to RMIS regions by overlaying the FRAM stock origin river with the RMIS atlas.

```{r add lipid ranking tu abundance db, message=FALSE}
lipid_ranking_params <-read_xlsx(here("R","data","SRKW_prey_kcal.xlsx"), sheet = 12) #Parameters to calculate the lipid content relative to lipid ranking and length

lipid_ranking_stock <- read_xlsx(here("R","data","SRKW_prey_kcal.xlsx"), sheet = 9) #Tab 'Lipid calls clean'
lipid_ranking_stock <- lipid_ranking_stock%>%distinct(FRAM.long.names,RMIS.Region, .keep_all = TRUE) #Collapsing by FRAM Stock and RMIS regions since the tier does not change among age classes.
lipid_ranking_stock <- lipid_ranking_stock[-6] #Remove 'Age' column

#Lipid rankings were assigned only to 'Marked' individuals. However, as the abundance database contains information on both marked and unmarked stocks,
#we assume that marked and unmarked animals from the same stock belong to the same lipid category.

lipid_ranking_stock_unmarked <- lipid_ranking_stock%>%mutate(FRAM.long.names = paste0("Un",FRAM.long.names))
lipid_ranking_stock <- rbind(lipid_ranking_stock,lipid_ranking_stock_unmarked)
lipid_ranking_stock$main.id <- c(1:length(lipid_ranking_stock$main.id)) #Re-generate main.id.

stock_abundance <- read_csv(here("R","data","Cohort_Stock_Shelton_seasons.csv"))

stock_abundance <- left_join(stock_abundance,lipid_ranking_stock,  by = c('StockLongName'='FRAM.long.names'))
```

The following stocks did not have lipid tier and were assigned to the 'medium' category:

_"Unmarked Nooksack Spr Natural"_
_"Marked Nooksack Spr Hatchery,"_ and
_"Unmarked Nooksack Spr Hatchery,"_ which belong to the NOWA RMIS region. 

Additionally, the stocks _"Unmarked Mid Oregon Coast Fall"_ and _"Marked Mid Oregon Coast Fall"_ were also assigned to the 'medium' lipid tier and are associated with the NOOR RMIS region.

```{r assigning lipid tiers and rmis regions to missing values}

stock_abundance <- stock_abundance%>%
  mutate(Lipid.Ranking.fix = if_else(is.na(`Lipid Ranking`),'medium',`Lipid Ranking`),
         RMIS.Region.fix = case_when(
      StockLongName %in% c("UnMarked Nooksack Spr Natural", "Marked Nooksack Spr Hatchery", "UnMarked Nooksack Spr Hatchery") ~ "NOWA",
      StockLongName %in% c("UnMarked Mid Oregon Coast Fall", "Marked Mid Oregon Coast Fall") ~ "NOOR",
      TRUE ~ RMIS.Region),
      
      FRAM.names.fix = str_remove(StockLongName, "^(Marked |UnMarked )"))

```
### Calculating Cohorts

The 'Cohort' year is necessary since the size estimates database needs to be linked to the FRAM abundance database. This way, it is possible to calculate the total lipids per stock/ocean age/year.

The assumption here is that the Cohort year (or Run year) is calculated as:

$$\text{Cohort} = \text{Brood year} + \text{Freshwater residence} + \text{Ocean age}$$
The `freshwater age` is estimated as follows: fish returning in spring (March-June) are 2 years old in freshwater (FW age-2), whereas fish returning in summer and fall (July-November) are 1 year old in freshwater (FW age-1).

We could derive the _Brood year_ as:

$$\text{Brood year}\ = \text{Cohort} -  \text{Freshwater residence} - \text{Ocean age}$$

```{r fw age}
stock_fw_age <- read_xlsx(here("R","data","FRAM stock_season run and fw age.xlsx"), sheet = 1)

stock_fw_age2 <- stock_fw_age%>%mutate(FRAM.long.names = paste0("Un",FRAM.long.names)) #The catalog has only 'Marked' stocks.

stock_fw_age_full <- rbind(stock_fw_age,stock_fw_age2)
stock_fw_age_full <- stock_fw_age_full[c(1,3)]

stock_abundance <-stock_abundance%>%
  left_join(stock_fw_age_full, by = c("StockLongName"="FRAM.long.names"))

sum(is.na(stock_abundance$Freshwater.age)) #3487

#Identify stocks missing Freshwater age
stock.missing.fw.age <- stock_abundance %>%
  filter(is.na(Freshwater.age)) %>%
  distinct(StockLongName)

#Add Freshwater age to stocks
stock.missing.fw.age <- stock.missing.fw.age %>%
  mutate(Freshwater.age = case_when(
    str_detect(StockLongName, "Fall") ~ 1, 
    str_detect(StockLongName, "Spr") ~ 2,
    TRUE ~ NA_real_  # Si no es Fall ni Spr, asigna NA
  ))

#colnames(stock.missing.fw.age)[1] <- "StockLongName"

#Add the new info to the abundance database
stock_abundance <- stock_abundance%>%
  left_join(stock.missing.fw.age, by = "StockLongName") %>%  # Link tables by StockLongName
  mutate(Freshwater.age = if_else(is.na(Freshwater.age.x), Freshwater.age.y, Freshwater.age.x))%>% 
  select(-Freshwater.age.y, -Freshwater.age.x)

stock_abundance <- stock_abundance%>%
  mutate(Brood.year = Year.run - Freshwater.age - Age)
```

Now the abundance and the size predictions databases are linked by 'RMIS region', 'brood year' and 'ocean age'

```{r link abundance and size predictions}
temp <- stock_abundance%>%
  select(StockID, Age, StartCohort,Shelton.TimeStep,Year.run, Season.run, StockLongName,StockName = FRAM.names.fix, RMIS.Region.fix, Lipid.Ranking.fix,Brood.year)%>%
  left_join(full_gam_predictions, by = c("RMIS.Region.fix"='region','Brood.year'='brood_year','Age'='ocean_age'))
```

SPS has abundance data but not size estimates. Therefore, it is assumed that the lengths of SPS salmon are similar and follow trends comparable to those from nearby areas, such as HOOD. 
```{r check link abundance and size predictions, echo = FALSE}
check1 <- data.frame(region = c(unique(full_gam_predictions$region)))
check2 <- data.frame(region = c(unique(stock_abundance$RMIS.Region.fix)))

anti_join(check2,check1, by = 'region')
```

```{r add HOOD data to SPS}
sps_values <- full_gam_predictions %>%
  filter(region == "HOOD")%>%
  mutate(region = "SPS")

# Step 2: Fill missing SPS values with HOOD values
stock_abundance <- temp %>%
  left_join(sps_values, by = c('RMIS.Region.fix'='region',"Brood.year"='brood_year', "Age"='ocean_age'), suffix = c("", ".sps")) %>% # Merge SPS's values
  mutate(
    fit = if_else(RMIS.Region.fix == "SPS" & is.na(fit), fit.sps, fit),
    se.fit = if_else(RMIS.Region.fix == "SPS" & is.na(se.fit), se.fit.sps, se.fit),
    ref_mean = if_else(RMIS.Region.fix == "SPS" & is.na(ref_mean), ref_mean.sps, ref_mean),
    rel_pred = if_else(RMIS.Region.fix == "SPS" & is.na(rel_pred), rel_pred.sps, rel_pred)
  ) %>%
  select(-ends_with(".sps")) # Remove temporary columns

rm(temp)
```

Since the abundance data span up to 2020, the size estimates from 2011 to 2017 (Brood year) are assumed to remain constant after 2010 which is the last year with size estimations. An example of this is shown in the plot below:
```{r filling gaps, fig.width=12, fig.height=24}

stock_abundance_full <-stock_abundance%>%
  arrange(StockLongName,Season.run,Age)%>%
  group_by(StockLongName,Season.run,Age)%>%
  fill(fit, se.fit, ref_mean, rel_pred, .direction = 'down')%>%
  ungroup()%>%
  arrange(StockLongName, Year.run, Shelton.TimeStep, Age)

ggplot(stock_abundance_full[stock_abundance_full$Season.run == '.Spr',], aes(x = Year.run, y = fit, color = as.factor(Age)))+
  geom_line()+
  facet_wrap(~StockLongName, ncol = 4)+
  theme_minimal()
```

### Calculate lipid content

Three lipid indexes were calculated:

`lipid_content_f` refers to lipids calculated from predicted lengths (`fit`),

`lipid_content_t` refers to lipids calculated from the reference mean lengths (`ref_mean`).

`lipid_content_c` refers to lipids calculated assuming that the lengths of the fish have not changed relative to the first brood year (`contant_length`).

The indexes were calculated according to the lipid ranking and the length-lipid relationship parameters found in _O'neill et al., 2014_:

For 'high´lipid ranking $$\text{kcal}^{-1} = 1.8034e^{-05}*(\text{length})^{3.0796}$$
For 'medium' lipid ranking $$\text{kcal}^{-1} = 1.1051e^{-05}*(\text{length})^{3.122}$$
For 'low' lipid ranking $$\text{kcal}^{-1} = 7.2074e^{-06}*(\text{length})^{3.143}$$


Then, the TOTAL lipid content (predicted, theoretical and constant) was calculated by multiplying the lipid context index by the abundance.
```{r calculating lipids}

constantlengthref <- stock_abundance_full %>%
  select(StockLongName,Year.run, Age, fit)%>%
  rename(constant_length = fit)%>%
  distinct(StockLongName, Year.run,Age, .keep_all = TRUE)%>%
  group_by(StockLongName,Age)%>%
  filter(Year.run == min(Year.run))%>%
  ungroup()%>%
  select(-Year.run)

stock_abundance_full <- stock_abundance_full%>%
  left_join(constantlengthref, by = c('StockLongName','Age'))%>%
  mutate(lipid_content_f = case_when(Lipid.Ranking.fix == "high" ~ lipid_ranking_params$a[1]*fit^(lipid_ranking_params$b[1]),
                                   Lipid.Ranking.fix == "medium" ~ lipid_ranking_params$a[2]*fit^(lipid_ranking_params$b[2]),
                                   Lipid.Ranking.fix == "low" ~ lipid_ranking_params$a[3]*fit^(lipid_ranking_params$b[3])),
         
         lipid_content_t = case_when(Lipid.Ranking.fix == "high" ~ lipid_ranking_params$a[1]*ref_mean^(lipid_ranking_params$b[1]),
                                   Lipid.Ranking.fix == "medium" ~ lipid_ranking_params$a[2]*ref_mean^(lipid_ranking_params$b[2]),
                                   Lipid.Ranking.fix == "low" ~ lipid_ranking_params$a[3]*ref_mean^(lipid_ranking_params$b[3])),
         
         lipid_content_c = case_when(Lipid.Ranking.fix == "high" ~ lipid_ranking_params$a[1]*constant_length^(lipid_ranking_params$b[1]),
                                   Lipid.Ranking.fix == "medium" ~ lipid_ranking_params$a[2]*constant_length^(lipid_ranking_params$b[2]),
                                   Lipid.Ranking.fix == "low" ~ lipid_ranking_params$a[3]*constant_length^(lipid_ranking_params$b[3])),
         
         total_lipid_content_f = lipid_content_f*StartCohort,
         total_lipid_content_t = lipid_content_t*StartCohort,
         total_lipid_content_c = lipid_content_c*StartCohort
         )

```

The following plot shows the comparison between the lipid content calculated from the predicted ( _solid line_ ) and the constant lengths ( _dashed line_ ) over time. Some lines appear cut because the size estimates were linked based on the availability of abundance data over time.

```{r total lipid content contant vs fit plot, fig.width=12, fig.height=36}
ggplot(stock_abundance_full[stock_abundance_full$Season.run == '.Spr' & stock_abundance_full$Age >3 ,],aes(x=Year.run, color = as.factor(Age)))+
  geom_line(aes(y=lipid_content_f), linetype  = "solid", size =0.7)+
  geom_line(aes(y=lipid_content_c), linetype = "dashed", size=0.7)+
  theme_minimal()+
  labs(title = "Predicted and Constant (Red) lipid content over time")+
  facet_wrap(~StockLongName, ncol = 5)
```

The plot shows the Total_lipid_content `predicted` (Solid line) and `constant` (Dashed line) over time for the different FRAM stocks. Each stock lipid content includes the Marked and UnMarked salmon abundance.

```{r total lipid content contant over time, fig.width=12, fig.height=36}
temp <- stock_abundance_full%>%
  group_by(StockName, Season.run, Year.run, Age)%>%
  summarise(sum_total_lipid_content_f = sum(total_lipid_content_f),
            sum_total_lipid_content_c = sum(total_lipid_content_c))%>%
  ungroup()


ggplot(temp[temp$Season.run == '.Spr' & temp$Age >3,], aes (x = Year.run, color = as.factor(Age)))+
  geom_line(aes( y = sum_total_lipid_content_f), linetype = 'solid', size = 0.7)+
  geom_line(aes( y = sum_total_lipid_content_c), linetype = 'dashed', size = 0.7)+
  theme_minimal()+
  facet_wrap(~StockName, ncol = 3, scales = 'free')
```


The plot show the Total_lipid_content `predicted` (Solid line) and `constant` (Dashed line) over time for **all stocks** by season.

```{r total lipid content contant vs fit plot 2, fig.width=12, fig.height=6}
temp <- stock_abundance_full%>%
  group_by(Season.run, Year.run)%>%
  summarise(sum_total_lipid_content_f = sum(total_lipid_content_f),
            sum_total_lipid_content_c = sum(total_lipid_content_c))%>%
  ungroup()

ggplot(temp, aes (x = Year.run, color = as.factor(Season.run)))+
  geom_line(aes( y = sum_total_lipid_content_f), linetype = 'solid', size = 0.5)+
  geom_line(aes( y = sum_total_lipid_content_c), linetype = 'dashed', size = 0.5)+
  labs(y = 'Total lipid content', x = 'Year run', title = 'Total lipid content over time for each season  - Dashed: Constant, Solid: Predicted')+
  theme_minimal()
```

### Linking data to Shelton's stock regions

Finally, the Shelton stocks are linked to the FRAM stocks. The sum of lipids is calculated by _Shelton Stock/Cohort/Season/Age_ level.

```{r add Shelton info, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=25, fig.height=25}
SheltonRef <- read.csv(here("R","data","FRAM Stock Names_v4.csv"))
SheltonRef <- SheltonRef[,c(2,4)]


stock_abundance_full <- stock_abundance_full%>%
  left_join(SheltonRef, by = c('StockLongName' = 'Stock.Long.Name'))%>%
  relocate(Origin.Shelton = Origin.Code.SHELTON,.after = RMIS.Region.fix)

temp <- stock_abundance_full%>%
  group_by(Origin.Shelton,Year.run,Season.run,Age)%>%
  summarise(Shl_total_lipid_content_f = sum(total_lipid_content_f), #Shl = Shelton level
            Shl_total_lipid_content_c = sum(total_lipid_content_c),
            Shl_total_lipid_content_t = sum(total_lipid_content_t))%>%
  ungroup()
```

The following plot shows the Lipid content over time (Only Spring season) - Dashed lines: lipids calculated using constant length, Solid lines: lipids calculated using predicted length".

```{r Shelton info plot, echo=FALSE,warning=FALSE, message=FALSE, fig.align='center', fig.width=25, fig.height=25}
ggplot(temp[temp$Season.run == '.Spr',], aes(x=Year.run, color = as.factor(Age)))+
  geom_line(aes(y=Shl_total_lipid_content_f), size =0.8)+
  geom_line(aes(y=Shl_total_lipid_content_c), size=0.8, linetype = 'dashed')+
  facet_wrap(~Origin.Shelton, nrow = 6, scales = "free_y")+
  labs(title = "Lipid content over time - Spring season - Dashed lines: lipids calculated using constant length, Solid lines: lipids calculated using predicted length", x= "Year run", y = "Total lipid content")+
  theme_minimal()#+
  #theme(axis.text.x = element_text(vjust = 0.5, hjust=1),
  #      strip.text = element_text(size = 18),
  #      axis.title = element_text(size=18),
  #      axis.text = element_text(size=18),
  #      legend.text = element_text(size=18),
  #      legend.title = element_text(size=18))

```

```{r export database only lipids for spring and calculate z-values, echo=FALSE, message=FALSE}
#library(janitor)

lipids.4.model <- temp%>%
  filter(Season.run == '.Spr')%>%
  select(Origin.Shelton,Year.run,Shl_total_lipid_content_f, Shl_total_lipid_content_c)%>%
  group_by(Origin.Shelton,Year.run)%>%
  summarise(Shl_total_lipid_content_f = sum(Shl_total_lipid_content_f),
            Shl_total_lipid_content_c = sum(Shl_total_lipid_content_c))%>%#NOTE: This calculation includes ALL ages (2-5)
  ungroup()%>%
  pivot_longer(cols = c(Shl_total_lipid_content_f,Shl_total_lipid_content_c), names_to = 'lipid_type', values_to = 'lipid_value')%>%
  pivot_wider(names_from = c(Origin.Shelton,lipid_type),values_from = lipid_value)%>%
  mutate(across(c(2:25), ~(.-mean(.))/sd(.)))%>% #Calculate Z-scores
  rename_with( ~str_replace(.x, "_Shl_total_lipid_content", ""), .cols = -Year.run)

#save(lipids.4.model, file = here("R","output","lipids4model.rda"))

#colnames(lipids.4.model)[5] <- 'Puso.N'
#colnames(lipids.4.model)[6] <- 'Puso.S'
#colnames(lipids.4.model)[8] <- 'Sgeo.S'
#colnames(lipids.4.model)[2:12] <- paste0('lip.',names(lipids.4.model[2:12]),'.z')

```


